---
title: "taxavlifestyle"
author: "Erica Ryu"
date: "6/21/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 10. Taxa vs lifestyle

The purpose is to identify associations between specific lifestyle factors and the DA taxa identified from the trend test. 

## load packages
```{r}
library(phyloseq)
library(data.table)
library(magrittr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(purrr)
library(heatmaply)
library(rstatix)
```

## load data
```{r}
phyloseq_complete <- readRDS("output/ps_complete.rds")

# subset based on extraction kit
qiagen <- subset_samples(phyloseq_complete, Condition == "Qiagen")
```

## colors
```{r}
twocolors <- c("deepskyblue", "lightsalmon" )
fourcolors <- c("darkslateblue", "deepskyblue", "lightblue3", "lightsalmon")
heatmap_col <- c("#009194", "white")
```

## prep taxa table
```{r}
phy_genus <-  tax_glom(qiagen, "Genus", NArm = FALSE);phy_genus
phy_genus_RA <- transform_sample_counts(phy_genus, function(otu) otu/sum(otu))
otu_table <- as.data.frame(as.matrix(phy_genus_RA@otu_table))

setDT(otu_table, keep.rownames = TRUE)[]
colnames(otu_table)[1] <- "SampleID"

otu_table2 <- otu_table %>% data.frame %>% set_rownames(.$SampleID)
otu_table2 <- otu_table2[,-1]
otu_table2 <- t(otu_table2)
tax_table <- as.data.frame(as.matrix(phy_genus_RA@tax_table))

otu_tax <- merge(otu_table2, tax_table, by = 0, all = TRUE)

# change colnames to genus name
otu_tax$Genus <- replace_na(otu_tax$Genus, "_unclassified")
# and add higher taxonomic level to unclassified
otu_tax$Phylum <- ifelse(is.na(otu_tax$Phylum), otu_tax$Kingdom, otu_tax$Phylum)
otu_tax$Class <- ifelse(is.na(otu_tax$Class), otu_tax$Phylum, otu_tax$Class)
otu_tax$Order <- ifelse(is.na(otu_tax$Order), otu_tax$Class, otu_tax$Order)
otu_tax$Family <- ifelse(is.na(otu_tax$Family), otu_tax$Order, otu_tax$Family)

otu_tax$Genus <- ifelse(otu_tax$Genus == "_unclassified", paste(otu_tax$Family, otu_tax$Genus, sep = ""), otu_tax$Genus)

# rename NA_classified
otu_tax$Genus <- gsub("NA_unclassified", "Unclassified", otu_tax$Genus)

# make genus rownames
rownames(otu_tax) <- otu_tax$Genus
# and remove taxonomic columns
otu_tax <- subset(otu_tax, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus))

# remove europeans because no survey data for Europeans
otu_tax_wo_euro <- dplyr::select(otu_tax, -contains("EUR"))

# change to dataframe
otu_tax <- as.data.frame(otu_tax_wo_euro)

# add rownames
rownames(otu_tax) <- rownames(otu_tax_wo_euro)
```

## subset to DA taxa
```{r}
# subset to taxa of interest
taxa <- c("Streptobacillus", "Porphyromonadaceae_unclassified", "Granulicatella", "Moraxella", "Simonsiella", "Neisseria", "Bacteroidetes_unclassified", "Brachymonas", "Atopobium")

# extracting data frame rows
selected_otu_tax <- otu_tax[rownames(otu_tax) %in% taxa, ] 
```

## set up metadata 
```{r}
meta <- as.data.frame(qiagen@sam_data)
df_meta <- as.data.frame(as.matrix(meta))
df_meta_num <- dplyr::select(df_meta, "SEX2":"RHR")

dim(df_meta_num)

## remove factors and samples with lots of NAs
# filter out Americans
df_meta_num$Lifestyle <- df_meta$Lifestyle
wo_euro <- subset(df_meta_num, Lifestyle != "Industrial")

# remove columns specific to women (pregnancy, menstruation, etc)
wo_mens <- dplyr::select(wo_euro, -c("MENS2":"MISC2"))

# remove columns for bitter taste perception
wo_bitter <- dplyr::select(wo_mens, -c("BTP_CTRL2":"BTP_SBZ2"))

# remove columns for vitals
wo_vital <- dplyr::select(wo_bitter, -c("HGT2":"RHR"))

# remove geographical info
wo_geo <- dplyr::select(wo_vital, -c("LAT2":"ALT2"))

## change sporadic NAs
# change NAs in EXER_FREQ2
wo_geo$EXER_FREQ2[is.na(wo_geo$EXER_FREQ2)] <- 0

# change NAs in SICK_LOC2
wo_geo$SICK_LOC2[is.na(wo_geo$SICK_LOC2)] <- 0

# change NAs in BRUSH_FREQ2
wo_geo$BRUSH_FREQ2[is.na(wo_geo$BRUSH_FREQ2)] <- 0

# change NAs in BLKT2
blkt <- wo_geo
blkt$BLKT2[is.na(blkt$BLKT2)] <- 0

# change NA in HS
household <- blkt
household$HS2[is.na(household$HS2)] <- 2

# final df for CA
df_meta_wo_lifestyle <- dplyr::select(household, -c("Lifestyle"))
df_meta_final <- as.data.frame(sapply(df_meta_wo_lifestyle, as.factor))

dim(df_meta_final)
```

## generate linear models
```{r}
# set up df for entry
lifestylevtaxa <- data.frame(matrix(ncol = 10, nrow = 37))
colnames(lifestylevtaxa) <- c("lifestyle", "streptobacillus", "porphyromonadaceae", "granulicatella", "moraxella", "simonsiella", "neisseria", "bacteroidetes_unclassified", "brachymonas", "atopobium")
lifestylevtaxa$lifestyle <- colnames(df_meta_final)

# streptobacillus
strepto <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Streptobacillus", ]))

strepto_vector <- c()

set.seed(123)
for(i in colnames(df_meta_final)){
  obj <- print(lm(as.formula(paste0("strepto ~ ", i)), data = df_meta_final))
  strepto_vector = c(strepto_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$streptobacillus <- strepto_vector

# porphyromonadaceae
porph <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Porphyromonadaceae_unclassified", ]))

porph_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("porph ~", i)),  data=df_meta_final))
  porph_vector = c(porph_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$porphyromonadaceae <- porph_vector

# granulicatella
granul <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Granulicatella", ]))

granul_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("granul ~", i)),  data=df_meta_final))
  granul_vector = c(granul_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$granulicatella <- granul_vector

# moraxella
morax <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Moraxella", ]))

morax_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("morax ~", i)),  data=df_meta_final))
  morax_vector = c(morax_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$moraxella <- morax_vector

# simonsiella
simon <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Simonsiella", ]))

simon_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("simon ~", i)),  data=df_meta_final))
  simon_vector = c(simon_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$simonsiella <- simon_vector

# neisseria
neiss <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Neisseria", ]))

neiss_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("neiss ~", i)),  data=df_meta_final))
  neiss_vector = c(neiss_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$neisseria <- neiss_vector

# bacteroidetes_unclassified
bacter <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Bacteroidetes_unclassified", ]))

bacter_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("bacter ~", i)),  data=df_meta_final))
  bacter_vector = c(bacter_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$bacteroidetes_unclassified <- bacter_vector

# brachymonas
brachy <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Brachymonas", ]))

brachy_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("brachy ~", i)),  data=df_meta_final))
  brachy_vector = c(brachy_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$brachymonas <- brachy_vector

# atopobium
atopo <- as.numeric(as.vector(selected_otu_tax[rownames(selected_otu_tax) %in% "Atopobium", ]))

atopo_vector <- c()

set.seed(123)
for(i in colnames(df_meta_wo_lifestyle)){
  obj <- print(lm(as.formula(paste0("atopo ~", i)),  data=df_meta_final))
  atopo_vector = c(atopo_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

lifestylevtaxa$atopobium <- atopo_vector
```

## multiple test adjustment
```{r}
lifestylevtaxaBH <- lifestylevtaxa

# strepto
lifestylevtaxaBH$strepto_BHadj_p_value <- p.adjust(lifestylevtaxaBH$streptobacillus, method = "BH", n = length(lifestylevtaxaBH$streptobacillus))

# porph
lifestylevtaxaBH$porph_BHadj_p_value <- p.adjust(lifestylevtaxaBH$porphyromonadaceae, method = "BH", n = length(lifestylevtaxaBH$porphyromonadaceae))

# granul
lifestylevtaxaBH$granul_BHadj_p_value <- p.adjust(lifestylevtaxaBH$granulicatella, method = "BH", n = length(lifestylevtaxaBH$granulicatella))

# morax
lifestylevtaxaBH$morax_BHadj_p_value <- p.adjust(lifestylevtaxaBH$moraxella, method = "BH", n = length(lifestylevtaxaBH$moraxella))

# simon
lifestylevtaxaBH$simon_BHadj_p_value <- p.adjust(lifestylevtaxaBH$simonsiella, method = "BH", n = length(lifestylevtaxaBH$simonsiella))

# neiss
lifestylevtaxaBH$neiss_BHadj_p_value <- p.adjust(lifestylevtaxaBH$neisseria, method = "BH", n = length(lifestylevtaxaBH$neisseria))

# bacter
lifestylevtaxaBH$bacter_BHadj_p_value <- p.adjust(lifestylevtaxaBH$bacteroidetes_unclassified, method = "BH", n = length(lifestylevtaxaBH$bacteroidetes_unclassified))

# brachy
lifestylevtaxaBH$brachy_BHadj_p_value <- p.adjust(lifestylevtaxaBH$brachymonas, method = "BH", n = length(lifestylevtaxaBH$brachymonas))

# atopo
lifestylevtaxaBH$atopo_BHadj_p_value <- p.adjust(lifestylevtaxaBH$atopobium, method = "BH", n = length(lifestylevtaxaBH$atopobium))

write.csv(lifestylevtaxaBH, file = "output/lifestylevtaxaBH.csv")
```

## individual microbes vs lifestyle factors
```{r}
# log transform
mk_logit <- function(otu_tax) {
  return(log10(otu_tax+0.0001))}
log_otu_tax <- map_df(otu_tax, .f = mk_logit)

rownames(log_otu_tax) <- rownames(otu_tax)
relabund_otu_tax <- t(log_otu_tax)
relabund_otu_tax <- as.data.frame(relabund_otu_tax)

# grain - add lifestyle info and plot
relabund_otu_tax_grain <- relabund_otu_tax
relabund_otu_tax_grain$grain <- as.character(df_meta_final$GRA2)
relabund_otu_tax_grain$grain <- ifelse(grepl("0",relabund_otu_tax_grain$grain), "Barley/Maize", "Rice/Wheat")
relabund_otu_tax_grain$grain <- factor(relabund_otu_tax_grain$grain, levels=c("Barley/Maize", "Rice/Wheat"))

# plot
plot_brachy_grain <- ggplot(relabund_otu_tax_grain, aes(x = grain, y = Brachymonas)) +
  geom_violin(aes(fill=grain), alpha=0.8) +
  geom_boxplot(width=0.1, color="black", alpha=0.2, outlier.shape = NA) +
  scale_x_discrete(limits=c("Barley/Maize", "Rice/Wheat")) +
  scale_fill_manual(name=NULL,
                    values=twocolors,
                    breaks=c("Barley/Maize", "Rice/Wheat"),
                    labels=c("Barley/Maize", "Rice/Wheat")) +
  geom_jitter(size = 1, col="darkgreen", width = 0.1) + 
  labs(title = "Brachymonas against Grain", x = "Grain", y = "log10 Relative Abundance")+
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust=0.5)) + 
  theme(legend.position = "none")

ggsave(file="figures/plot_brachy_grain.pdf", plot=plot_brachy_grain, width=3, height=4)

# smoking - add lifestyle info and plot
relabund_otu_tax_smok <- relabund_otu_tax
relabund_otu_tax_smok$smoking <- as.character(df_meta_final$SMOK2)
relabund_otu_tax_smok$smoking <- ifelse(grepl("0",relabund_otu_tax_smok$smoking), "No", "Yes")
relabund_otu_tax_smok$smoking <- factor(relabund_otu_tax_smok$smoking, levels=c("Yes", "No"))

smok_taxa <- relabund_otu_tax_smok[,c("Porphyromonadaceae_unclassified","Granulicatella", "Neisseria","Atopobium", "smoking")]

smok_taxa <- subset(relabund_otu_tax_smok, select = c("Porphyromonadaceae_unclassified","Granulicatella", "Neisseria","Atopobium", "smoking"))

# change format
smok_taxa <- pivot_longer(smok_taxa, !smoking, names_to = "Taxa", values_to = "Rel_abund")

# set level
smok_taxa$smoking <- factor(smok_taxa$smoking, ordered = TRUE, levels=c("Yes", "No"))

plot_smok <- ggplot(smok_taxa, aes(x = smoking, y = Rel_abund, group = smoking), color = black) +
  geom_violin(aes(fill=smoking), alpha=0.8) +
  geom_boxplot(width=0.1, color="black", alpha=0.2, outlier.shape = NA) +
  scale_x_discrete(limits=c("Yes", "No")) +
  scale_fill_manual(name=NULL,
                    values=twocolors,
                    breaks=c("Yes", "No"),
                    labels=c("Yes", "No")) +
  geom_jitter(size = 1, width=0.1, col="darkgreen") + facet_wrap(~Taxa, ncol = 2, scales = "fixed") +
  scale_y_continuous(labels = function(x) format(x, nsmall = 1)) +
  labs(title = "Taxa against Smoking", x = "Smoking", y = "log10 Relative Abundance")+
  theme(strip.text = element_text(face = "italic")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  theme(legend.position = "none")

ggsave(file="figures/plot_smok.pdf", plot=plot_smok, width=6, height=4)

# sisnu
# subset
relabund_otu_tax_sisnu <- relabund_otu_tax
## sisnu
relabund_otu_tax_sisnu$Sisnu <- as.character(df_meta_final$Sisnu2)
relabund_otu_tax_sisnu$Sisnu <- ifelse(grepl("0",relabund_otu_tax_sisnu$Sisnu), "No", "Yes")
relabund_otu_tax_sisnu$Sisnu <- factor(relabund_otu_tax_sisnu$Sisnu, levels=c("Yes", "No"))

sisnu_taxa <- relabund_otu_tax_sisnu[,c("Porphyromonadaceae_unclassified","Brachymonas", "Moraxella", "Sisnu")]

# change format
sisnu_taxa <- pivot_longer(sisnu_taxa, !Sisnu, names_to = "Taxa", values_to = "Rel_abund")

# set level
sisnu_taxa$Sisnu <- factor(sisnu_taxa$Sisnu, ordered = TRUE, levels=c("Yes", "No"))

plot_sisnu <- ggplot(sisnu_taxa, aes(x = Sisnu, y = Rel_abund, group = Sisnu), color = black) +
  geom_violin(aes(fill=Sisnu), alpha=0.8) +
  geom_boxplot(width=0.1, color="black", alpha=0.2, outlier.shape = NA) +
  scale_x_discrete(limits=c("Yes", "No")) +
  scale_fill_manual(name=NULL,
                    values=twocolors,
                    breaks=c("Yes", "No"),
                    labels=c("Yes", "No")) +
  geom_jitter(size = 1, width=0.15, col="darkgreen") + facet_wrap(~Taxa, ncol = 2, scales = "fixed") +
  labs(title = "Taxa against Sisnu", x = "Sisnu consumption", y = "log10 Relative Abundance")+
  theme(strip.text = element_text(face = "italic")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  theme(legend.position = "none")

ggsave(file="figures/plot_sisnu.pdf", plot=plot_sisnu, width=6, height=4)
```

## heatmap of lifestyle v taxa
```{r}
lifestylevtaxa_hm <- lifestylevtaxa

# change to human readable labels
rownames(lifestylevtaxa_hm) <- c("sex", "literacy", "location", "drinking_water", "fuel", "kitchen_loc", "toilet", "grain", "Sisnu", "fish", "meat", "black_tea", "milk_tea", "soda", "milk", "yogurt", "yogurt_freq", "fermented", "ferm_freq", "food_source", "scarcity", "smoking", "tobacco", "alcohol", "exercise", "exercise_freq", "sick_checkup", "checkup_loc", "health_travel", "meds", "ayurvedic", "brushing", "brushing_freq", "age", "education", "household", "children")

lifestylevtaxa_hm <- subset(lifestylevtaxa_hm, select = -c(lifestyle))

lifestylevtaxa_hm <- as.matrix(lifestylevtaxa_hm)

# plot and save
hm_plot <- heatmaply(lifestylevtaxa_hm,
          colors = heatmap_col, file = "figures/hm_plot.pdf", width = 700, height = 800, text_angle = 15)

# NOTE: if heatmaply fails to save the heatmap, either run webshot::install_phantomjs() or install orca at https://github.com/plotly/orca#installation
```

## distribution of smokers and alcohol drinkers
```{r}
lifestyle_distrib <- household

lifestyle_distrib_factors <- subset(lifestyle_distrib, select = c(SMOK2, ALCO2, Lifestyle))

lifestyle_distrib_factors$Lifestyle <- gsub("RecentlySettled", "Recently Settled", lifestyle_distrib_factors$Lifestyle)

lifestyle_distrib_factors$SMOK2 <- ifelse(grepl("0",lifestyle_distrib_factors$SMOK2), "No", "Yes")
lifestyle_distrib_factors$ALCO2 <- ifelse(grepl("0",lifestyle_distrib_factors$ALCO2), "No", "Yes")

smok_count <- lifestyle_distrib_factors %>% count(Lifestyle,SMOK2)
alc_count <- lifestyle_distrib_factors %>% count(Lifestyle,ALCO2)

smok_count$SMOK2 <- factor(smok_count$SMOK2, levels=c("Yes", "No"))
alc_count$ALCO2 <- factor(alc_count$ALCO2, levels=c("Yes", "No"))

smok_count$Lifestyle <- factor(smok_count$Lifestyle, levels=c("Foragers", "Recently Settled", "Agriculturalists", "Expats"))
alc_count$Lifestyle <- factor(alc_count$Lifestyle, levels=c("Foragers", "Recently Settled", "Agriculturalists", "Expats"))
 
# Stacked
smok_distrib <- ggplot(smok_count, aes(fill=SMOK2, y=n, x=Lifestyle)) + 
    geom_bar(position="fill", stat="identity") +
    scale_fill_manual(name=NULL,
                    values=twocolors,
                    breaks=c("Yes", "No"),
                    labels=c("Yes", "No")) +
  labs(title = "Smoking distribution",
       x = "Lifestyle",
       y = "Proportion")

ggsave(file="figures/smok_distrib.pdf", plot=smok_distrib, width=6, height=2)

alco_distrib <- ggplot(alc_count, aes(fill=ALCO2, y=n, x=Lifestyle)) + 
    geom_bar(position="fill", stat="identity") +
    scale_fill_manual(name=NULL,
                    values=twocolors,
                    breaks=c("Yes", "No"),
                    labels=c("Yes", "No")) +
  labs(title = "Alcohol distribution",
       x = "Lifestyle",
       y = "Proportion")

ggsave(file="figures/alco_distrib.pdf", plot=alco_distrib, width=6, height=2)

# test for association
## smoking
smok_conting <- as.data.frame(smok_count %>% pivot_wider(names_from = SMOK2, values_from = n))

smok_conting$Lifestyle  <- ifelse(smok_conting$Lifestyle == "Foragers", "1.Foragers",
                              ifelse(smok_conting$Lifestyle == "Recently Settled", "2.Recently Settled",
                              ifelse(smok_conting$Lifestyle == "Agriculturalists", "3.Agriculturalists","4.Expats")))

rownames(smok_conting) <- smok_conting$Lifestyle
smok_conting <- subset(smok_conting, select = -c(Lifestyle))
smok_conting <- smok_conting[order(row.names(smok_conting)), , drop = FALSE]

smok_chisq <- prop_trend_test(smok_conting)

## alcohol
alc_conting <- as.data.frame(alc_count %>% pivot_wider(names_from = ALCO2, values_from = n))

alc_conting$Lifestyle  <- ifelse(alc_conting$Lifestyle == "Foragers", "1.Foragers",
                              ifelse(alc_conting$Lifestyle == "Recently Settled", "2.Recently Settled",
                              ifelse(alc_conting$Lifestyle == "Agriculturalists", "3.Agriculturalists","4.Expats")))

rownames(alc_conting) <- alc_conting$Lifestyle
alc_conting <- subset(alc_conting, select = -c(Lifestyle))
alc_conting <- alc_conting[order(row.names(alc_conting)), , drop = FALSE]

alc_chisq <- prop_trend_test(alc_conting)
```

