---
title: "network_analysis"
author: "Erica Ryu"
date: "6/6/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 14. Network Analysis

The purpose of this script is to generate a network of the oral microbiome

## install SpiecEasi (if not already installed)
```{r}
# library(devtools)
# install_github("zdk123/SpiecEasi")
# library(SpiecEasi)
```

## load packages
```{r}
library(phyloseq)
library(SpiecEasi)
library(Matrix)
library(data.table)
library(dplyr)
library(tidyr)
library(igraph)
library(ggplot2)
library(magrittr)
library(boot)
```

## load data
```{r}
# oral microbiome phyloseq
phyloseq <- readRDS("output/ps_complete.rds")

# subset based on extraction kit
qiagen <- subset_samples(phyloseq, Condition == "Qiagen")
```

## prep oral microbiome data
```{r}
phy_genus <- tax_glom(qiagen, "Genus", NArm = FALSE);phy_genus
otu_table <- as.data.frame(as.matrix(phy_genus@otu_table))

setDT(otu_table, keep.rownames = TRUE)[]
colnames(otu_table)[1] <- "SampleID"

otu_table2 <- otu_table %>% data.frame %>% magrittr::set_rownames(.$SampleID)
otu_table2 <- otu_table2[,-1]
otu_table2 <- t(otu_table2)
tax_table <- as.data.frame(as.matrix(phy_genus@tax_table))

otu_tax <- merge(otu_table2, tax_table, by = 0, all = TRUE)

# change colnames to genus name
otu_tax$Genus <- replace_na(otu_tax$Genus, "_unclassified")
# and add higher taxonomic level to unclassified
otu_tax$Phylum <- ifelse(is.na(otu_tax$Phylum), otu_tax$Kingdom, otu_tax$Phylum)
otu_tax$Class <- ifelse(is.na(otu_tax$Class), otu_tax$Phylum, otu_tax$Class)
otu_tax$Order <- ifelse(is.na(otu_tax$Order), otu_tax$Class, otu_tax$Order)
otu_tax$Family <- ifelse(is.na(otu_tax$Family), otu_tax$Order, otu_tax$Family)

otu_tax$Genus <- ifelse(otu_tax$Genus == "_unclassified", paste(otu_tax$Family, otu_tax$Genus, sep = ""), otu_tax$Genus)

# rename completely unclassified ASV
otu_tax$Genus <- gsub("NA_unclassified", "Unclassified", otu_tax$Genus)

# make genus rownames
rownames(otu_tax) <- otu_tax$Genus
# and remove taxonomic columns
otu_tax <- subset(otu_tax, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus))

otu_tax_t <- t(otu_tax)
```

## run sparcc
```{r}
set.seed(100)
sparcc_output <- sparcc(otu_tax_t)
```

## format sparcc outputs
```{r}
# Define threshold for SparCC correlation matrix for the graph
sparcc.graph <- abs(sparcc_output$Cor) >= 0.5
diag(sparcc.graph) <- 0
sparcc.graph <- Matrix(sparcc.graph, sparse=TRUE)
# Create igraph object
ig.sparcc <- adj2igraph(sparcc.graph, vertex.attr=list(name=rownames(otu_tax)))
# apply layout
am.coord <- layout.fruchterman.reingold(ig.sparcc)
```

## centrality
```{r}
degr_cent_sparcc <- centr_degree(ig.sparcc, mode = 'all')
degr_cent_sparcc <- degr_cent_sparcc$res

# Compute betweenness centrality
betw_cent_sparcc <- igraph::betweenness(ig.sparcc)

# compute closeness centrality
close_cent_sparcc <- igraph::closeness(ig.sparcc)

# make into one dataframe
sparcc_cent <- data.frame(label = rownames(otu_tax),
                   degree = degr_cent_sparcc, 
                   betweenness = betw_cent_sparcc,
                   closeness = close_cent_sparcc)

# plot histogram of degree distributions
ggplot(data = sparcc_cent, aes(x = degree)) +
  geom_bar(alpha = .5, position = 'identity') +
  ggtitle('Degree distribution')
```

## examine taxa with at least one edge
```{r}
# plot taxa with at least one edge
isolated_sparcc <- which(degree(ig.sparcc)==0)
ig.sparcc_wo0 <- delete.vertices(ig.sparcc, isolated_sparcc)
am.coord_sparcc <- am.coord[-isolated_sparcc,]
plot(ig.sparcc_wo0, 
    vertex.size = 5+degr_cent_sparcc,   # Change node size
     vertex.shape = 'circle',      # Specify node shape
     asp = 0,  layout = am.coord_sparcc)

# get centrality only for taxa with edges
sparcc_cent_wo0 <- filter(sparcc_cent, degree != 0)
```

## calculate modularity and make CAGs
```{r}
set.seed(100)
modul_sparcc <- cluster_edge_betweenness(ig.sparcc_wo0)
modularity(modul_sparcc)

modul_sparcc_CAGs <- modul_sparcc$membership

clusterOneIndices_sparcc <- which(modul_sparcc$membership==1)
clusterOneOtus_sparcc <- modul_sparcc$names[clusterOneIndices_sparcc]
clusterTwoIndices_sparcc <- which(modul_sparcc$membership==2)
clusterTwoOtus_sparcc <- modul_sparcc$names[clusterTwoIndices_sparcc]
clusterThreeIndices_sparcc <- which(modul_sparcc$membership==3)
clusterThreeOtus_sparcc <- modul_sparcc$names[clusterThreeIndices_sparcc]
clusterFourIndices_sparcc <- which(modul_sparcc$membership==4)
clusterFourOtus_sparcc <- modul_sparcc$names[clusterFourIndices_sparcc]
clusterFiveIndices_sparcc <- which(modul_sparcc$membership==5)
clusterFiveOtus_sparcc <- modul_sparcc$names[clusterFiveIndices_sparcc]
clusterSixIndices_sparcc <- which(modul_sparcc$membership==6)
clusterSixOtus_sparcc <- modul_sparcc$names[clusterSixIndices_sparcc]

# combine into one dataframe
CAGs <- stack(mget(c("clusterOneOtus_sparcc", "clusterTwoOtus_sparcc", "clusterThreeOtus_sparcc", "clusterFourOtus_sparcc", "clusterFiveOtus_sparcc", "clusterSixOtus_sparcc")))
```

## save tables for input into cytoscope for visualization
```{r}
write.graph(ig.sparcc_wo0, file="output/sparcc_v3.txt", format="ncol")
write.table(CAGs, file="output/CAGs_v3.txt", sep="\t", row.names = FALSE, quote=FALSE)
```

## ratio of CAGs across groups
```{r}
# subset genera table to each CAG
CAG_gen <- subset(otu_tax_t, select = c(clusterOneOtus_sparcc, clusterTwoOtus_sparcc, clusterThreeOtus_sparcc, clusterFourOtus_sparcc, clusterFiveOtus_sparcc, clusterSixOtus_sparcc))

CAG_gen <- as.data.frame(CAG_gen)

# reformat CAGs into relative proportions
## separate into individual CAGs
CAG1_gen <- subset(CAG_gen, select = c(clusterOneOtus_sparcc))
CAG2_gen <- subset(CAG_gen, select = c(clusterTwoOtus_sparcc))
CAG3_gen <- subset(CAG_gen, select = c(clusterThreeOtus_sparcc))
CAG4_gen <- subset(CAG_gen, select = c(clusterFourOtus_sparcc))
CAG5_gen <- subset(CAG_gen, select = c(clusterFiveOtus_sparcc))
CAG6_gen <- subset(CAG_gen, select = c(clusterSixOtus_sparcc))

# find total for each CAG for each sample
CAG1 <- rowSums(CAG1_gen)
CAG2 <- rowSums(CAG2_gen)
CAG3 <- rowSums(CAG3_gen)
CAG4 <- rowSums(CAG4_gen)
CAG5 <- rowSums(CAG5_gen)
CAG6 <- rowSums(CAG6_gen)
CAG_total <- rowSums(CAG_gen)

# make all values relative to total within sample
CAG_proportions <- cbind(CAG1, CAG2, CAG3, CAG4, CAG5, CAG6, CAG_total)
CAG_proportions <- CAG_proportions/CAG_proportions[,"CAG_total"]

# add lifestyle and reformat
CAG_lifestyle <- CAG_proportions
CAG_lifestyle <- subset(CAG_lifestyle, select = -c(CAG_total))
CAG_lifestyle <- as.data.frame(CAG_lifestyle)
CAG_lifestyle$lifestyle <- ifelse(grepl("CHE", rownames(CAG_lifestyle)), "Foragers", ifelse(grepl("EUR", rownames(CAG_lifestyle)), "American Industrialist", ifelse(grepl("NEW00", rownames(CAG_lifestyle)) | grepl("NEW10", rownames(CAG_lifestyle)) | grepl("THA", rownames(CAG_lifestyle)), "Agriculturalists", ifelse(grepl("NEW01", rownames(CAG_lifestyle)) | grepl("NEW11", rownames(CAG_lifestyle)), "Expats", "Recently Settled"))))

# get mean value for each lifestyle per CAG
CAG_forg <- c(mean(CAG_lifestyle$CAG1[CAG_lifestyle$lifestyle == "Foragers"]),
              mean(CAG_lifestyle$CAG2[CAG_lifestyle$lifestyle == "Foragers"]),
              mean(CAG_lifestyle$CAG3[CAG_lifestyle$lifestyle == "Foragers"]),
              mean(CAG_lifestyle$CAG4[CAG_lifestyle$lifestyle == "Foragers"]),
              mean(CAG_lifestyle$CAG5[CAG_lifestyle$lifestyle == "Foragers"]),
              mean(CAG_lifestyle$CAG6[CAG_lifestyle$lifestyle == "Foragers"]))
CAG_rs <- c(mean(CAG_lifestyle$CAG1[CAG_lifestyle$lifestyle == "Recently Settled"]),
            mean(CAG_lifestyle$CAG2[CAG_lifestyle$lifestyle == "Recently Settled"]),
            mean(CAG_lifestyle$CAG3[CAG_lifestyle$lifestyle == "Recently Settled"]),
            mean(CAG_lifestyle$CAG4[CAG_lifestyle$lifestyle == "Recently Settled"]),
            mean(CAG_lifestyle$CAG5[CAG_lifestyle$lifestyle == "Recently Settled"]),
            mean(CAG_lifestyle$CAG6[CAG_lifestyle$lifestyle == "Recently Settled"]))
CAG_ag <- c(mean(CAG_lifestyle$CAG1[CAG_lifestyle$lifestyle == "Agriculturalists"]),
            mean(CAG_lifestyle$CAG2[CAG_lifestyle$lifestyle == "Agriculturalists"]),
            mean(CAG_lifestyle$CAG3[CAG_lifestyle$lifestyle == "Agriculturalists"]),
            mean(CAG_lifestyle$CAG4[CAG_lifestyle$lifestyle == "Agriculturalists"]),
            mean(CAG_lifestyle$CAG5[CAG_lifestyle$lifestyle == "Agriculturalists"]),
            mean(CAG_lifestyle$CAG6[CAG_lifestyle$lifestyle == "Agriculturalists"]))
CAG_expat <- c(mean(CAG_lifestyle$CAG1[CAG_lifestyle$lifestyle == "Expats"]),
               mean(CAG_lifestyle$CAG2[CAG_lifestyle$lifestyle == "Expats"]),
               mean(CAG_lifestyle$CAG3[CAG_lifestyle$lifestyle == "Expats"]),
               mean(CAG_lifestyle$CAG4[CAG_lifestyle$lifestyle == "Expats"]),
               mean(CAG_lifestyle$CAG5[CAG_lifestyle$lifestyle == "Expats"]),
               mean(CAG_lifestyle$CAG6[CAG_lifestyle$lifestyle == "Expats"]))
CAG_indus <- c(mean(CAG_lifestyle$CAG1[CAG_lifestyle$lifestyle == "American Industrialist"]),
               mean(CAG_lifestyle$CAG2[CAG_lifestyle$lifestyle == "American Industrialist"]),
               mean(CAG_lifestyle$CAG3[CAG_lifestyle$lifestyle == "American Industrialist"]),
               mean(CAG_lifestyle$CAG4[CAG_lifestyle$lifestyle == "American Industrialist"]),
               mean(CAG_lifestyle$CAG5[CAG_lifestyle$lifestyle == "American Industrialist"]),
               mean(CAG_lifestyle$CAG6[CAG_lifestyle$lifestyle == "American Industrialist"]))

# combine into one table
CAG_avg <- as.data.frame(rbind(CAG_forg, CAG_rs, CAG_ag, CAG_expat, CAG_indus))
colnames(CAG_avg) <- c("CAG1", "CAG2", "CAG3", "CAG4", "CAG5", "CAG6")
CAG_avg$lifestyle <- c("Foragers", "Recently Settled", "Agriculturalists", "Expats", "American Industrialist")

# reformat for plotting
CAG_lifestyle_longer <- pivot_longer(CAG_avg, cols=c("CAG1", "CAG2", "CAG3", "CAG4", "CAG5", "CAG6"), names_to = "CAG", values_to = "proportion")

# add levels
CAG_lifestyle_longer$lifestyle <- factor(CAG_lifestyle_longer$lifestyle, levels=c("Foragers", "Recently Settled", "Agriculturalists", "Expats", "American Industrialist"))

CAG_colors <- c("#FFFF00", "#FF0033", "#FFFFFF", "#0000FF", "#66CCFF", "#000000")

CAG_pie <- ggplot(CAG_lifestyle_longer, aes(x="", y=proportion, group=CAG, fill=CAG)) +
  geom_bar(width = 1, stat = "identity", color = "dark gray", linewidth = 0.2) + 
  scale_fill_manual(name=NULL,
                    values=CAG_colors,
                    breaks=c("CAG1", "CAG2", "CAG3", "CAG4", "CAG5", "CAG6"),
                    labels=c("CAG1", "CAG2", "CAG3", "CAG4", "CAG5", "CAG6")) +
         coord_polar("y", start=0) +  
         facet_grid(.~ lifestyle) + 
         guides(fill = FALSE) +
         theme_void() 

ggsave(file="figures/CAG_pie.pdf", plot=CAG_pie, width=8, height=3)

```