---
title: "decontam"
author: "Erica Ryu"
date: "12/6/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2. Remove contaminants via decontam

## The purpose of this script is to remove contaminants via decontam using both the frequency and prevalance methods

## load packages
```{r}
library(phyloseq)
library(ggplot2)
library(decontam)
library(microbiome)
library(dplyr)
library(data.table)
```

## load in phyloseq object
```{r}
phyloseq <- readRDS("output/oralmicrobiome_ps.rds")
```

## check library sizes
```{r}
df <- as.data.frame(sample_data(phyloseq)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(phyloseq)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
df$Sample_or_Control <- ifelse(is.na(df$Lifestyle), "control", "sample" )

# plot
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + 
  geom_point()
```

## remove contaminating taxa using "frequency" method
### In this method, the distribution of the frequency of each sequence feature as a function of the input DNA concentration is used to identify contaminants.
```{r}
# remove samples that say yes or NA in the BDNA column
phyloseq_BDNA <-  subset_samples(phyloseq, !BDNA=="Yes")
phyloseq_BDNA <-  subset_samples(phyloseq_BDNA, !BDNA=="NA")

# format data for decontam
phyloseq_BDNA@sam_data$BDNA <- as.numeric(phyloseq_BDNA@sam_data$BDNA)
contamdf.freq <- isContaminant(phyloseq_BDNA, method="frequency", conc="BDNA")
head(contamdf.freq) # The default value of threshold = 0.1 is used, and $contaminant=TRUE if $p < 0.1

# number of contaminants
table(contamdf.freq$contaminant)

# how abundant are the contaminant ASVs?
head(which(contamdf.freq$contaminant))

# plot contaminant ASV vs noncontaminant and save as pdf
png("output/freq_plot.pdf")
plot_frequency(phyloseq_BDNA, taxa_names(phyloseq_BDNA)[c(2,865)], conc="BDNA") + 
  xlab("DNA Concentration")
dev.off() 
```

## remove frequency contaminant taxa from phyloseq object
```{r}
phyloseq.noncontam <- prune_taxa(!contamdf.freq$contaminant, phyloseq)
```

## identify contaminants with the prevalence method
### This method compares the prevalence of a sequence feature in true positive samples to its prevalence in negative controls
```{r}
## add sample_or_control column
sample_data(phyloseq.noncontam)$Sample_or_Control <- ifelse(is.na(phyloseq.noncontam@sam_data$Lifestyle), "control", "sample")
sample_data(phyloseq.noncontam)$is.neg <- sample_data(phyloseq.noncontam)$Sample_or_Control == "control"
contamdf.prev <- isContaminant(phyloseq.noncontam, method="prevalence", neg="is.neg")

# how many contaminants?
table(contamdf.prev$contaminant)

# how abundant are the contaminant ASVs?
which(contamdf.prev$contaminant)

# how many times were taxa observed in neg controls vs pos samples?
phyloseq.pa <- transform_sample_counts(phyloseq.noncontam, function(abund) 1*(abund>0))
phyloseq.pa.neg <- prune_samples(sample_data(phyloseq.pa)$Sample_or_Control == "control", phyloseq.pa)
phyloseq.pa.pos <- prune_samples(sample_data(phyloseq.pa)$Sample_or_Control == "sample", phyloseq.pa)

# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(phyloseq.pa.pos), pa.neg=taxa_sums(phyloseq.pa.neg),
                      contaminant=contamdf.prev$contaminant)

# plot
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
```

## remove contaminant taxa
```{r}
phyloseq.noncontam_all <- prune_taxa(!contamdf.prev$contaminant, phyloseq.noncontam)

## save phyloseq object with contaminants removed
saveRDS(phyloseq.noncontam_all, "output/ps_noncontam.rds")
```

## examine which taxa are getting removed
```{r}
# frequency method
tax <- as.data.frame(as.matrix(phyloseq@tax_table))
contaminant_taxa_freq <- filter(contamdf.freq, contaminant == TRUE) # filter to just contaminants
contam_tax_freq <- merge(contaminant_taxa_freq, tax, by = "row.names")

# prevalence method
contaminant_taxa_prev <- filter(contamdf.prev, contaminant == TRUE) # filter to just contaminants
contam_tax_prev <- merge(contaminant_taxa_prev, tax, by = "row.names")
```

## and in which samples the taxa are found
```{r}
# combine all contaminant taxa IDs into one df
contam_tax_freq$decon <- rep(c("freq"), 6)
contam_tax_prev$decon <- rep(c("prev"), 13)
contam_tax_comb <- rbind(contam_tax_freq, contam_tax_prev)
contam_tax_comb <- as.data.frame(as.matrix(contam_tax_comb))

# subset OTU table based on contaminant taxa 
otu <- as.data.frame(as.matrix(otu_table(phyloseq)))
otu_transpose <- transpose(otu)
colnames(otu_transpose) <- rownames(otu)
Row.names <- colnames(otu)
otu_transpose <- cbind(Row.names, otu_transpose)
contam_tax_samp <- subset(otu_transpose, Row.names %in% contam_tax_comb$Row.names)

# separate OTU table by extraction kit
qiagen_contam_tax <- select(contam_tax_samp, contains("genotek"))
psoil_contam_tax <- select(contam_tax_samp, contains("psoil"))

# remove samples with no contaminating taxa
qiagen_contam_tax <- qiagen_contam_tax[, colSums(qiagen_contam_tax!= 0) > 0]
psoil_contam_tax <- psoil_contam_tax[, colSums(psoil_contam_tax!= 0) > 0]

# check 
dim(qiagen_contam_tax)
dim(psoil_contam_tax)

# remove samples with less than 10 counts of contaminating taxa
qiagen_contam_tax_10 <- qiagen_contam_tax[, colSums(qiagen_contam_tax) > 10]
dim(qiagen_contam_tax_10)
psoil_contam_tax_10 <- psoil_contam_tax[, colSums(psoil_contam_tax) > 10]
dim(psoil_contam_tax_10)

# make taxID as row name
qiagen_contam_tax_10$Row.names <- contam_tax_samp$Row.names
psoil_contam_tax_10$Row.names <- contam_tax_samp$Row.names

# merge with taxonomic classification
qiagen_taxID <- merge(contam_tax_comb, qiagen_contam_tax_10, by = "Row.names")
psoil_taxID <- merge(contam_tax_comb, psoil_contam_tax_10, by = "Row.names")

# remove taxa that 
qiagen_taxID  <- qiagen_taxID[rowSums(qiagen_taxID[grep("_genotek", colnames(qiagen_taxID))]) > 0, ]
psoil_taxID  <- psoil_taxID[rowSums(psoil_taxID[grep("_psoil", colnames(psoil_taxID))]) > 0, ]

# save tables listing which contaminants are primarily in which samples
write.csv(qiagen_taxID, file = "output/qiagen_contaminants.csv")
write.csv(psoil_taxID, file = "output/psoil_contaminants.csv")
```