---
title: "aldex2_diff_abund"
author: "Erica Ryu"
date: "3/16/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 7. Differential abundance

The purpose of this script is to conduct difference abundance analysis with ALDEx2.

## load packages
```{r}
library(phyloseq)
library(data.table)
library(dplyr)
library(magrittr)
library(tidyr)
library(ALDEx2)
```

## load phyloseq object and set seed
```{r}
set.seed(123)
# oral microbiome phyloseq
phyloseq_complete <- readRDS("output/ps_complete.rds")

# subset based on extraction kit
qiagen <- subset_samples(phyloseq_complete, Condition == "Qiagen")

# gut microbiome phyloseq
gut_phyloseq <- readRDS("data/gut_phyloseq.rds")
```

## load colors
```{r}
fivecolors <- c("darkslateblue", "deepskyblue", "lightblue3", "lightsalmon" , "firebrick")
```

## prep taxa table for ALDEx2
```{r}
phy_genus <-  tax_glom(qiagen, "Genus", NArm = FALSE)
otu_table <- as.data.frame(as.matrix(phy_genus@otu_table))

setDT(otu_table, keep.rownames = TRUE)[]
colnames(otu_table)[1] <- "SampleID"

otu_table2 <- otu_table %>% data.frame %>% set_rownames(.$SampleID)
otu_table2 <- otu_table2[,-1]
otu_table2 <- t(otu_table2)
tax_table <- as.data.frame(as.matrix(phy_genus@tax_table))

otu_tax <- merge(otu_table2, tax_table, by = 0, all = TRUE)

# change colnames to genus name
otu_tax$Genus <- replace_na(otu_tax$Genus, "_unclassified")

# add higher taxonomic level name to unclassified
otu_tax$Phylum <- ifelse(is.na(otu_tax$Phylum), otu_tax$Kingdom, otu_tax$Phylum)
otu_tax$Class <- ifelse(is.na(otu_tax$Class), otu_tax$Phylum, otu_tax$Class)
otu_tax$Order <- ifelse(is.na(otu_tax$Order), otu_tax$Class, otu_tax$Order)
otu_tax$Family <- ifelse(is.na(otu_tax$Family), otu_tax$Order, otu_tax$Family)

otu_tax$Genus <- ifelse(otu_tax$Genus == "_unclassified", paste(otu_tax$Family, otu_tax$Genus, sep = ""), otu_tax$Genus)

# rename completely unclassified ASV
otu_tax$Genus <- gsub("NA_unclassified", "Unclassified", otu_tax$Genus)

# set genus column as rownames
rownames(otu_tax) <- otu_tax$Genus

# remove taxonomic columns
otu_tax <- subset(otu_tax, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus))
```

## ALDEx2 kw
```{r}
# set up comparison groups
conds <- ifelse(grepl("CHE",otu_table$SampleID), "Foragers", 
                    ifelse(grepl("EUR", otu_table$SampleID), "American Industrialist",
                              ifelse(grepl("NEW00",otu_table$SampleID) | grepl("NEW10",otu_table$SampleID) | grepl("THA",otu_table$SampleID), 
                                     "Agriculturalists",
                              ifelse(grepl("NEW01",otu_table$SampleID) | grepl("NEW11",otu_table$SampleID), "Expats", "Recently Settled"))))

# transform data
set.seed(123)
transform_kw <- aldex.clr(otu_tax, conds, mc.samples=1000, denom="all")

# generate model
model_kw <- aldex.kw(transform_kw)
```

## prep gut microbiome taxa table
```{r}
phy_genus_gut <-  tax_glom(gut_phyloseq, "Genus", NArm = FALSE);phy_genus_gut
otu_table_gut <- as.data.frame(as.matrix(phy_genus_gut@otu_table))

setDT(otu_table_gut, keep.rownames = TRUE)[]
colnames(otu_table_gut)[1] <- "SampleID"

otu_table2_gut <- otu_table_gut %>% data.frame %>% set_rownames(.$SampleID)
otu_table2_gut <- otu_table2_gut[,-1]
otu_table2_gut <- t(otu_table2_gut)
tax_table_gut <- as.data.frame(as.matrix(phy_genus_gut@tax_table))

otu_tax_gut <- merge(otu_table2_gut, tax_table_gut, by = 0, all = TRUE)

# change colnames to genus name
otu_tax_gut$Genus <- replace_na(otu_tax_gut$Genus, "_unclassified")
# and add higher taxonomic level to unclassified
otu_tax_gut$Phylum <- ifelse(is.na(otu_tax_gut$Phylum), otu_tax_gut$Kingdom, otu_tax_gut$Phylum)
otu_tax_gut$Class <- ifelse(is.na(otu_tax_gut$Class), otu_tax_gut$Phylum, otu_tax_gut$Class)
otu_tax_gut$Order <- ifelse(is.na(otu_tax_gut$Order), otu_tax_gut$Class, otu_tax_gut$Order)
otu_tax_gut$Family <- ifelse(is.na(otu_tax_gut$Family), otu_tax_gut$Order, otu_tax_gut$Family)

otu_tax_gut$Genus <- ifelse(otu_tax_gut$Genus == "_unclassified", paste(otu_tax_gut$Family, otu_tax_gut$Genus, sep = ""), otu_tax_gut$Genus)

# rename completely unclassified ASV
otu_tax_gut$Genus <- gsub("NA_unclassified", "Unclassified", otu_tax_gut$Genus)

# make genus rownames
rownames(otu_tax_gut) <- otu_tax_gut$Genus
# and remove taxonomic columns
otu_tax_gut <- subset(otu_tax_gut, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus))
```

## aldex kw for gut microbiome
```{r}
# set up comparison groups
conds_gut <- ifelse(grepl("CHE",otu_table_gut$SampleID), "Foragers", 
                ifelse(grepl("EUR",otu_table_gut$SampleID), "American Industrialist",
                              ifelse(grepl("THA",otu_table_gut$SampleID), "Agriculturalists", "Recently Settled")))

# transform data
set.seed(100)
transform_kw_gut <- aldex.clr(otu_tax_gut, conds_gut, mc.samples=1000, denom="all")

# generate model
model_kw_gut <- aldex.kw(transform_kw_gut)
```

## only keep methods that are being analyzed
```{r}
model_kw <- subset(model_kw, select = c("kw.ep", "kw.eBH"))
model_kw_gut <- subset(model_kw_gut, select = c("kw.ep", "kw.eBH"))
```

## compare oral and gut
```{r}
model_kw_filt <- model_kw[model_kw$kw.eBH < 0.05,]
model_kw_gut_filt <- model_kw_gut[model_kw_gut$kw.eBH < 0.05,] 
```

## save ALDEx2 results
```{r}
write.csv(model_kw, file = "output/model_kw.csv")
write.csv(model_kw_gut, file = "output/model_kw_gut.csv")
```

## prep for comparison with added covariates
```{r}
no_am_ps <- subset_samples(qiagen, Lifestyle != "Industrial")

no_am_genus <-  tax_glom(no_am_ps, "Genus", NArm = FALSE)
otu_no_am <- as.data.frame(as.matrix(no_am_genus@otu_table))

setDT(otu_no_am, keep.rownames = TRUE)[]
colnames(otu_no_am)[1] <- "SampleID"

otu_no_am2 <- otu_no_am %>% data.frame %>% set_rownames(.$SampleID)
otu_no_am2 <- otu_no_am2[,-1]
otu_no_am2 <- t(otu_no_am2)
tax_no_am <- as.data.frame(as.matrix(no_am_genus@tax_table))

otu_tax_no_am <- merge(otu_no_am2, tax_no_am, by = 0, all = TRUE)

# change colnames to genus name
otu_tax_no_am$Genus <- replace_na(otu_tax_no_am$Genus, "_unclassified")

# add higher taxonomic level name to unclassified
otu_tax_no_am$Phylum <- ifelse(is.na(otu_tax_no_am$Phylum), otu_tax_no_am$Kingdom, otu_tax_no_am$Phylum)
otu_tax_no_am$Class <- ifelse(is.na(otu_tax_no_am$Class), otu_tax_no_am$Phylum, otu_tax_no_am$Class)
otu_tax_no_am$Order <- ifelse(is.na(otu_tax_no_am$Order), otu_tax_no_am$Class, otu_tax_no_am$Order)
otu_tax_no_am$Family <- ifelse(is.na(otu_tax_no_am$Family), otu_tax_no_am$Order, otu_tax_no_am$Family)

otu_tax_no_am$Genus <- ifelse(otu_tax_no_am$Genus == "_unclassified", paste(otu_tax_no_am$Family, otu_tax_no_am$Genus, sep = ""), otu_tax_no_am$Genus)

# rename completely unclassified ASV
otu_tax_no_am$Genus <- gsub("NA_unclassified", "Unclassified", otu_tax_no_am$Genus)

# set genus column as rownames
rownames(otu_tax_no_am) <- otu_tax_no_am$Genus

# remove taxonomic columns
otu_tax_no_am <- subset(otu_tax_no_am, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus))
```

## compare with added covariates
```{r}
# all samples
## covariates
sex <- qiagen@sam_data$SEX
lifestyle <- qiagen@sam_data$Lifestyle

lifestyle <- gsub("Foragers", "1.Foragers", lifestyle)
lifestyle <- gsub("RecentlySettled", "2.Recently Settled", lifestyle)
lifestyle <- gsub("Agriculturalists", "3.Agriculturalists", lifestyle)
lifestyle <- gsub("Expats", "4.Expats", lifestyle)
lifestyle <- gsub("Industrial", "5.American Industrialist", lifestyle)

## just lifestyle
covariates_lifestyle <- data.frame(lifestyle)
mm_lifestyle <- model.matrix(~ lifestyle, covariates_lifestyle)

set.seed(100)
x.glm <- aldex.clr(otu_tax, mm_lifestyle, mc.samples=1000, denom="all", verbose=T)
glm.test <- aldex.glm(x.glm, mm_lifestyle)

## lifestyle and sex
covariates_sex <- data.frame(lifestyle, sex)

mm_sex <- model.matrix(~ lifestyle + sex, covariates_sex)

set.seed(100)
x.glm.sex <- aldex.clr(otu_tax, mm_sex, mc.samples=1000, denom="all", verbose=T)
glm.test.sex <- aldex.glm(x.glm.sex, mm_sex)


# no americans
## covariates
sex_noam <- no_am_ps@sam_data$SEX
age <- no_am_ps@sam_data$AGE
lifestyle_noam <- no_am_ps@sam_data$Lifestyle

lifestyle_noam <- gsub("Foragers", "1.Foragers", lifestyle_noam)
lifestyle_noam <- gsub("RecentlySettled", "2.Recently Settled", lifestyle_noam)
lifestyle_noam <- gsub("Agriculturalists", "3.Agriculturalists", lifestyle_noam)
lifestyle_noam <- gsub("Expats", "4.Expats", lifestyle_noam)

## no americans
covariates_noam <- data.frame(lifestyle_noam)

mm_noam <- model.matrix(~ lifestyle_noam, covariates_noam)

set.seed(100)
x.glm.noam <- aldex.clr(otu_tax_no_am, mm_noam, mc.samples=1000, denom="all", verbose=T)
glm.test.noam <- aldex.glm(x.glm.noam, mm_noam)

## all covariates
covariates_all <- data.frame(age, lifestyle_noam)

mm_all <- model.matrix(~ age  + lifestyle_noam, covariates_all)

set.seed(100)
x.glm.all <- aldex.clr(otu_tax_no_am, mm_all, mc.samples=1000, denom="all", verbose=T)
glm.test.all <- aldex.glm(x.glm.all, mm_all)
```

## BH correction
```{r}
glm.test$`lifestyle2.Recently Settled:pval.BH` <- p.adjust(glm.test$`lifestyle2.Recently Settled:pval`, method = "BH", n = length(glm.test$`lifestyle2.Recently Settled:pval`))
glm.test$`lifestyle3.Agriculturalists:pval.BH` <- p.adjust(glm.test$`lifestyle3.Agriculturalists:pval`, method = "BH", n = length(glm.test$`lifestyle3.Agriculturalists:pval`))
glm.test$`lifestyle4.Expats:pval.BH` <- p.adjust(glm.test$`lifestyle4.Expats:pval`, method = "BH", n = length(glm.test$`lifestyle4.Expats:pval`))
glm.test$`lifestyle5.American Industrialist:pval.BH` <- p.adjust(glm.test$`lifestyle5.American Industrialist:pval`, method = "BH", n = length(glm.test$`lifestyle5.American Industrialist:pval`))

glm.test.sex$`lifestyle2.Recently Settled:pval.BH` <- p.adjust(glm.test.sex$`lifestyle2.Recently Settled:pval`, method = "BH", n = length(glm.test.sex$`lifestyle2.Recently Settled:pval`))
glm.test.sex$`lifestyle3.Agriculturalists:pval.BH` <- p.adjust(glm.test.sex$`lifestyle3.Agriculturalists:pval`, method = "BH", n = length(glm.test.sex$`lifestyle3.Agriculturalists:pval`))
glm.test.sex$`lifestyle4.Expats:pval.BH` <- p.adjust(glm.test.sex$`lifestyle4.Expats:pval`, method = "BH", n = length(glm.test.sex$`lifestyle4.Expats:pval`))
glm.test.sex$`lifestyle5.American Industrialist:pval.BH` <- p.adjust(glm.test.sex$`lifestyle5.American Industrialist:pval`, method = "BH", n = length(glm.test.sex$`lifestyle5.American Industrialist:pval`))

glm.test.noam$`lifestyle_noam2.Recently Settled:pval.BH` <- p.adjust(glm.test.noam$`lifestyle_noam2.Recently Settled:pval`, method = "BH", n = length(glm.test.noam$`lifestyle_noam2.Recently Settled:pval`))
glm.test.noam$`lifestyle_noam3.Agriculturalists:pval.BH` <- p.adjust(glm.test.noam$`lifestyle_noam3.Agriculturalists:pval`, method = "BH", n = length(glm.test.noam$`lifestyle_noam3.Agriculturalists:pval`))
glm.test.noam$`lifestyle_noam4.Expats:pval.BH` <- p.adjust(glm.test.noam$`lifestyle_noam4.Expats:pval`, method = "BH", n = length(glm.test.noam$`lifestyle_noam4.Expats:pval`))

glm.test.all$`lifestyle_noam2.Recently Settled:pval.BH` <- p.adjust(glm.test.all$`lifestyle_noam2.Recently Settled:pval`, method = "BH", n = length(glm.test.all$`lifestyle_noam2.Recently Settled:pval`))
glm.test.all$`lifestyle_noam3.Agriculturalists:pval.BH` <- p.adjust(glm.test.all$`lifestyle_noam3.Agriculturalists:pval`, method = "BH", n = length(glm.test.all$`lifestyle_noam3.Agriculturalists:pval`))
glm.test.all$`lifestyle_noam4.Expats:pval.BH` <- p.adjust(glm.test.all$`lifestyle_noam4.Expats:pval`, method = "BH", n = length(glm.test.all$`lifestyle_noam4.Expats:pval`))
```

## only keep pertinent columns
```{r}
glm.test.BH <- subset(glm.test, select = c("lifestyle2.Recently Settled:pval", "lifestyle3.Agriculturalists:pval", "lifestyle4.Expats:pval", "lifestyle5.American Industrialist:pval", "lifestyle2.Recently Settled:pval.BH", "lifestyle3.Agriculturalists:pval.BH", "lifestyle4.Expats:pval.BH", "lifestyle5.American Industrialist:pval.BH"))

glm.test.sex.BH <- subset(glm.test.sex, select = c("lifestyle2.Recently Settled:pval", "lifestyle3.Agriculturalists:pval", "lifestyle4.Expats:pval", "lifestyle5.American Industrialist:pval", "lifestyle2.Recently Settled:pval.BH", "lifestyle3.Agriculturalists:pval.BH", "lifestyle4.Expats:pval.BH", "lifestyle5.American Industrialist:pval.BH"))

glm.test.noam.BH <- subset(glm.test.noam, select = c("lifestyle_noam2.Recently Settled:pval", "lifestyle_noam3.Agriculturalists:pval", "lifestyle_noam4.Expats:pval", "lifestyle_noam2.Recently Settled:pval.BH", "lifestyle_noam3.Agriculturalists:pval.BH", "lifestyle_noam4.Expats:pval.BH"))

glm.test.all.BH <- subset(glm.test.all, select = c("lifestyle_noam2.Recently Settled:pval", "lifestyle_noam3.Agriculturalists:pval", "lifestyle_noam4.Expats:pval", "lifestyle_noam2.Recently Settled:pval.BH", "lifestyle_noam3.Agriculturalists:pval.BH", "lifestyle_noam4.Expats:pval.BH"))
```

## save ALDEx2 covariate comparison
```{r}
write.csv(glm.test.BH, file = "output/glm.test.BH.csv")
write.csv(glm.test.sex.BH, file = "output/glm.test.sex.BH.csv")
write.csv(glm.test.noam.BH, file = "output/glm.test.noam.BH.csv")
write.csv(glm.test.all.BH, file = "output/glm.test.all.BH.csv")
```
